---
author: Michael Raffanti
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      error = TRUE, fig.height = 3)
library(tidyverse)
library(lubridate)
source("../../scripts/viridis.R")
```

## Homework Assignment 4

### Problems

This assignment involves five data sets involving international flights arriving to Chicago's O'Hare airport from January 1, 2017 through December 31, 2021
with one separate file for each year.
Each file has been cleaned from the original HTML.

Each data set is in a separate CSV file:
`ORD-2017.csv`, `ORD-2018.csv`, `ORD-2019.csv`, `ORD-2020.csv`, `ORD-2021.csv`.

This code chunk reads in these five data sets and combines them into a single data set.

```{r}
ord = map_dfr(2017:2021, ~{
  return ( read_csv(str_c("../../data/ORD-", .x, ".csv")) )
})
```

### 1

Make the following changes to the data set `ord`.

- Add columns for *year*, *month* (character valued, use month abbreviations, Jan - Dec), *day* (day of the month), and *wday* (day of the week, character valued, use day abbreviations, Sun - Sat).  
- Reorder the variables so that these new variables all appear directly after the date column.  
- Remove the `terminal` variable.  
- Rename *all_total* to *passengers*, *all_flights* to *flights*, and *all_booths* to *booths*.  
- Arrange the rows by date and hour.  
- Print all columns of the first 5 rows of this modified data frame
- Print the dimensions of this modified data frame

```{r}
ord = ord %>% mutate(year = year(date), month = month(date, label=TRUE), day = day(date), wday = wday(date, label=TRUE))
ord = ord %>% select(airport, date, year, month, day, wday, everything()) %>% select(-(terminal))
ord = ord %>% rename(passengers = all_total, flights = all_flights, booths = all_booths)
ord = ord %>% arrange(date, hour)
head(ord, c(5,4))
ord
```



### 2

Are there any dates in the range from January 1, 2017 through December 31, 2021 that are missing? If so, which ones?

```{r}
temp = tibble(date = seq(mdy("January 1, 2017"), mdy("December 31, 2021"), 1))
temp %>% anti_join(ord, by = "date")
```

### 3

Modify the code chunk below to do the following tasks.

Add a variable named `time_of_day` to the `ord` data set which categorizes the `hour` variable in the following way:

- midnight to 4am ("0000 - 0100" through "0300 - 0400") is "overnight"
- 4am to 8am is "early morning"
- 8am to noon is "morning"
- noon to 4pm is "afternoon"
- 4pm to 8pm is "early evening"
- 8pm to midnight is "late evening"

Create a table which count the number of rows for each `time_of_day` and `hour` combination and print all rows of this table
to verify that you have created this variable correctly.

```{r}
## Add the variable and reorder its values
##   so that they follow the numerical hour variable
##   instead of being alphabetical
ord = ord %>% mutate(first_hour = as.numeric(str_sub(hour, 1, 4))) %>% mutate(time_of_day = case_when(0 <= first_hour & first_hour < 400 ~ 'overnight', 400 <= first_hour & first_hour < 800 ~ 'early morning', 800 <= first_hour & first_hour < 1200 ~ 'morning', 1200 <= first_hour & first_hour < 1600 ~ 'afternoon', 1600 <= first_hour & first_hour < 2000 ~ 'early evening', 2000 <= first_hour & first_hour < 2400 ~ 'late evening')) %>% mutate(time_of_day = reorder(time_of_day, first_hour)) %>% select(-(first_hour))

ord %>% 
  count(time_of_day, hour) %>% 
  print(n = Inf)
```

Calculate a summary table which calculates the total number of flights
which arrive during each of these time periods.

Print the table.

```{r}
prob3 = ord %>% group_by(time_of_day) %>% summarize(sum_flights = sum(flights)) 
prob3
```

Display these totals in a bar graph.

```{r}
ggplot(ord, aes(x = time_of_day, y = flights)) + geom_col() + xlab("Time of Day") + ylab("Number of Total Flights") + ggtitle("Number of Total Flights for each Time of Day")
```



### 4

- Create a data summary table with the average daily number of passengers by month and year.
- Display this data summary with a bar chart where month is on the x-axis and average daily passenger counts are on the y axis 
- Add meaningful axis labels and a title to this graph.  
- Change the y axis scale labels so values are regular numbers with commas. *(Use `scale_y_continuous()` as below)*
- Display the graph with a different facet for each year

```{r, fig.height = 6}
## modify this code block

## data summary
prob4 = ord %>% group_by(year, month) %>% summarize(passengers=mean(passengers))
prob4
  ## add code
  
## plot the table 
ggplot(prob4, aes(x = month, y = passengers)) + geom_col() + facet_grid(vars(year)) +
  scale_y_continuous(label = scales::comma) + xlab("Month") + ylab("Average Number of Passengers") + ggtitle("Average Number of Passengers for every Month, 2017-2021")
```


### 5

What striking feature is apparent in this graph?
What explains this feature of the data?

Obviously when we know COVID first appeared and lockdowns began flights were very rare, and we can see the extreme drop off of number of flights and passengers during that time. For the years 2017-2019, the peak for average number of passengers is in the summer months, which makes sense for those out of school and vacation from work. Flights started to increase after almost two years of the pandemic as seen in the end of 2021.

### 6

> Investigate the average number of passengers per day of the week.

Consider only the years 2017-2019.

- For each day of the week (Sunday - Saturday), calculate the average number of arriving daily passengers to ORD on international flights. Display this table sorted from highest to lowest average.
*(Hint: you need to total passengers by date and day of week before taking averages.)*
- Print the entire table

```{r}
prob6 = ord %>% filter(year >= 2017 & year <= 2019) %>% group_by(date, wday) %>% group_by(wday) %>% summarize(passengers = mean(passengers))
prob6
```


### 7

Identify the 20 dates from 2017-2019 with the highest total number of arriving passengers.

How many of these dates are the day of the week identified in the previous problem with the highest average?

```{r}
prob7 = ord %>% filter(year >= 2017 & year <= 2019) %>% group_by(date, wday) %>% summarize(passengers = mean(passengers)) %>% ungroup() %>% slice_max(passengers, n = 20)
prob7
```


